Paper 3 - Causal Roadmap

### 2023-03-08 docs/causal_roadmap.qmd
Finished first draft of roadmap.

---
title: "Paper 3 - Causal Roadmap"
keywords:
  - Causal Inference
  - Epidemiology
  - Triangulation
  - Modified Treatment Policy
  - Mendelian Randomization
  - Metabolomics
  - Endocrine Disrupting Chemicals
  - Neurodevelopment
  - Neurobehavior
author:
  - name: "Lorenzo Fabbri"
    number: 1
    email: lorenzo.fabbri@isglobal.org
    orcid: 0000-0003-3031-322X
    attributes:
      corresponding: true
    affiliations:
      - ref: isg
      - ref: upf
affiliations:
  - id: isg
    name: Barcelona Institute for Global Health (ISGlobal)
    city: Barcelona
    state: Spain
  - id: upf
    name: Pompeu Fabra University
    city: Barcelona
    state: Spain
format:
  pdf:
    pdf-engine: pdflatex
    number-sections: true
    toc: true
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 20mm
    margin-top: 15mm
    include-in-header:
      - \usepackage{rotating}
      - \usepackage{float}
    pandoc_args:
      - !expr acronymsdown::add_filter()
bibliography: ../../../PhD.bib
link-citations: true
linkcolor: blue
execute:
  echo: false
editor_options: 
  chunk_output_type: console
filters:
  - /home/lorenzo/R/x86_64-pc-linux-gnu-library/4.1/acronymsdown/parse-acronyms.lua
acronyms:
  insert_links: false
  sorting: alphabetical
  keys:
    - key: edc
      shortname: EDC
      longname: endocrine disruptor
    - key: edcs
      shortname: EDCs
      longname: endocrine disruptors
    - key: mtp
      shortname: MTP
      longname: modified treatment policy
    - key: mtps
      shortname: MTPs
      longname: modified treatment policies
    - key: tmle
      shortname: TMLE
      longname: Targeted Minimum Loss-Based Estimator
    - key: sl
      shortname: SL
      longname: Super Learner
    - key: cpm
      shortname: CPM
      longname: Raven’s Coloured Progressive Matrices
    - key: ant
      shortname: ANT
      longname: Attention Network Test
    - key: cbcl
      shortname: CBCL
      longname: Child Behavior Checklist
    - key: npsem
      shortname: NPSEM
      longname: non-parametric Structural Equation Model
    - key: prs
      shortname: PRS
      longname: polygenic risk score
    - key: mr
      shortname: MR
      longname: Mendelian Randomization
    - key: se
      shortname: SE
      longname: socio-economic
    - key: iv
      shortname: IV
      longname: instrumental variable
    - key: ivs
      shortname: IVs
      longname: instrumental variables
    - key: ctd
      shortname: CTD
      longname: Comparative Toxicogenomics Database
    - key: cv
      shortname: CV
      longname: cross-validation
---

@@

\textcolor{red}{Most of the following was **copy-pasted** from published papers.}

@@

The **aim** of the present study is to research the short-term effects of postnatal exposure to non-persistent \acr{edcs} on neurodevelopment and neurobehavior in childhood, and how the metabolome and the proteome might mediate these effects. We will strengthen the estimated effects by making use of the principles and criteria of triangulation [@LawlorTillingDaveySmith:2016]. To estimate these effects, we will primarily rely on the use of \acr{mtps} [@munozPopulationInterventionCausal2012; @haneuseEstimationEffectInterventions2013; @diazNonparametricCausalEffects2021] in combination with \acr{tmle} [@vanderlaanTargetedMinimumLossBased2018], to avoid relying on arbitrary parametric assumptions. We will consider both additive and multiplicative shifts of the exposures' distribution.

@@

\textcolor{red}{TODO}

@@

@@

where $(f_{W{_t}}, f_{A_{t}}, f_{Y_{t}})$ are the non-parametric structural equations, 
and $(U_{W_{t}}, U_{A_{t}}, U_{Y_{t}})$ are the unmeasured factors contributing the confounders, exposures, and outcome, respectively. 
The corresponding causal graph (without indexes for ease of notation) is given in @fig-causal-graph.

```{r}
#| label: fig-causal-graph
#| fig.cap: Causal graph corresponding to structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal_graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") %>%
  ggdag::tidy_dagitty()
ggdag::ggdag(causal_graph) + ggdag::theme_dag()
```

@@

## Research question 1: What are the short-term effects of childhood exposure to \acr{edcs} on the outcomes?

@@

## Research question 2: What are the short-term effects of childhood exposure to \acr{edcs} on metabolites and proteins?

We will estimate the short-term effects of childhood exposure to non-persistent \acr{edcs} on metabolites and proteins measured in childhood. We will perform an MWAS analysis with the dependent variable $m_i$ being the metabolite or protein $i$, and the independent variable $x_j$ being the levels of \acr{edc} $j$. We will assume that all associations are confounded by the same set of confounders, independently of the \acr{edc} considered.

## Research question 3: What are the short-term effects of childhood *exposure* to metabolites and proteins on the outcomes?

We will estimate the short-term effects of childhood *exposure* to metabolites and proteins on the outcomes of interest measured in childhood. We will perform an MWAS analysis with the dependent variable $y$ being the outcome, and the independent variable $m_i$ being the metabolite or protein $i$. We will assume that all associations are confounded by the same set of confounders, independently of the \acr{edc} considered.

@@

* **No unmeasured confounding**: There are no unmeasured common causes of the exposure and the subsequent outcome, which we can formalize as $Y^d \perp \!\!\! \perp A|W$. In our context, this assumption would be violated if, for example, an unmeasured variable influences both the exposure levels and the clinical outcome. We cannot guarantee that this assumption holds, and therefore limit our interpretations to statistical associations rather than causal effects.
* **Positivity**: If $(a, w)$ is within the support of $A, W$, then $(d(a,c), w)$ for additive shifts and $(d(a,k), w)$ for multiplicative shifts must also be within the support of $A,W$. In practice, this means that for any given time index and set of adjustment covariates, there is a positive probability of finding a subject with the same covariate values and a exposure level matching the shifted value. We will attempt to improve plausibility of this assumption by considering small shifts, while recognizing that this approach makes our causal effect data-adaptive.
* **Independence** of subjects. This assumption also implies **no interference**: the exposure level $a$ for a given subject does not affect the outcomes of the other subjects.
* **Consistency**: If $A = a$ for any subject, then $Y(a) = Y$, and hence the full observed set of outcomes when $A^d = A$ is simply $Y(A^d) = Y$. This means that the counterfactual outcome for a subject with its observed exposure level is the observed outcome.
* **Time-ordering**: The confounders $W$ precede the exposure $A$, which also precedes the outcome $Y$.

@@

with $\Lambda$ indicating whether each subject received the natural or shifted value of treatment. We refer to $P(\Lambda=1|a,w)$ as $\bar{\lambda}$. To implement estimation of $\bar{\lambda}$ and $\bar{Q}$ flexibly, we will use the ensemble machine learning algorithm \acr{sl}. In the following, we present the steps necessary to estimate $\psi_0 (A^d)$ and the associational parameter $\psi_0^{\Delta}$. We note that our bounded continuous outcome will be **scaled** to $[0,1]$, and that to handle missing covariates $W_{i,j}$, we will redefine the data to include $\delta_{ij}$, where $\delta_{ij}=1$ if the covariate was observed and $0$ otherwise.

@@

Using the empirical mean $\hat{Y} = \frac{1}{n} \sum_{i=1}^nY_i$, to estimate $\psi_0(A) = \mathbb{E}(Y)$, we obtain an estimate of the statistical association parameter $\psi_0^{\Delta} = \psi_0(A^d) = \psi_0(A)$ with
$$
\hat{\psi}^{\Delta} = \hat{\psi}_{\text{tmle}} - \bar{Y}.
$$

@@

Using `r I(cv.folds)`-fold \acr{cv}, \acr{sl} builds the best weighted combination of predictions from the algorithms listed in Table \@ref(tab:lmtp-specs). In case of repeated measures of the exposures, an `ID` variable will be used to make sure that the `r I(cv.folds)`-fold \acr{cv} splits will keep observations from the same individuals in the same split.

@@

- `outcome ~ chemical`
  - Regression analysis. We will use \acr{tmle} in combination with \acr{sl}, and a library of estimators including both simple parametric models (e.g., `glm`) and data-adaptive semi-parametric models. The models will be adjusted for relevant confounders, as identified with `dagitty`. We hypothesize the presence of residual confounding, especially due to genetic and parental \acr{se} factors, which would result in **exaggeration** of any true causal effect.
  - Outcome negative control study. We will try to identify outcomes in the same cohorts that would be affected by confounders that are relevant in the association of \acr{edcs} exposure to the clinical outcome of interest, but for which a biological/causal effect of the exposures is unlikely.
  - Literature search. For each of the `outcome ~ chemical` associations, we will perform a literature search to compare the obtained results. We will make use of PubMed, using as keywords the clinical outcome and the *common* name of each chemical, and the \acr{ctd} [@DavisWiegersJohnson:2023].
  - Qualitative \acr{prs}. \textcolor{red}{TODO}
- `omic ~ chemical`
  - Regression analysis. As above.
  - Cross-cohort comparison. We will compare the regression models' results between subjects of different ethnic origins. We postulate that \acr{se} factors represent the main confounders for this association. Based on the associations been \acr{se} factors (e.g., maternal education) and the omic markers, we will compare the results of `omic ~ chemical` with the expected modified associations (e.g., weaker or stronger).
  - Literature search. As above.
  - Qualitative \acr{prs}. \textcolor{red}{TODO}
- `outcome ~ omic`
  - Regression analysis. As above.
  - Cross-cohort comparison. As above.
  - \acr{mr}. We will make use, ideally, of a weighted allele score of genetic variants known to be robustly associated with the omic markers as an \acr{IV}. We will use methods, including sensitivity analysis, to explore the possibility of bias due to: (i) weak instruments, and (ii) violation of the exclusion restriction criteria. Based on these, we will postulate whether the obtained results might be biased towards the null or not. By using \acr{mr}, we assume that this approach tests *exposure* to omic markers across the whole of childhood.

## \acr{mr}: Assumptions

@@

@@

---

@@

---

@@

## Checklist A: Reproducibility

@@

### Research question 1:...

```{r}
#| fig.cap: DAG for Research question 1.
#| fig.height: 10.0
#| fig.width: 18.0
dag_rq1 <- dags$chem_to_out %>%
  dagitty::dagitty()
dag_rq1 %>% ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 10, text_col = "black") +
  ggdag::theme_dag() + ggdag::scale_adjusted() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))
```

The `r I(type_mas)` sufficient adjustment sets for estimating the `r I(type_effect)` effect of X on Y are:

```{r}
dagitty::adjustmentSets(x = dag_rq1, 
                        type = type_mas, effect = type_effect)
```

>>>>

---
title: "Paper 3 - Causal Roadmap"
keywords:
  - Causal Inference
  - Epidemiology
  - Triangulation
  - Modified Treatment Policy
  - Mendelian Randomization
  - Metabolomics
  - Endocrine Disrupting Chemicals
  - Neurodevelopment
  - Neurobehavior
author:
  - name: "Lorenzo Fabbri"
    number: 1
    email: lorenzo.fabbri@isglobal.org
    orcid: 0000-0003-3031-322X
    attributes:
      corresponding: true
    affiliations:
      - ref: isg
      - ref: upf
affiliations:
  - id: isg
    name: Barcelona Institute for Global Health (ISGlobal)
    city: Barcelona
    state: Spain
  - id: upf
    name: Pompeu Fabra University
    city: Barcelona
    state: Spain
format:
  pdf:
    pdf-engine: pdflatex
    number-sections: true
    toc: true
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 20mm
    margin-top: 15mm
    include-in-header:
      - \usepackage{rotating}
      - \usepackage{float}
      - \usepackage{typearea}
    pandoc_args:
      - !expr acronymsdown::add_filter()
bibliography: ../../../PhD.bib
link-citations: true
linkcolor: blue
execute:
  echo: false
editor_options: 
  chunk_output_type: console
filters:
  - /home/lorenzo/R/x86_64-pc-linux-gnu-library/4.1/acronymsdown/parse-acronyms.lua
acronyms:
  fromfile: ../../../all_phd/acronyms.yml
  insert_links: false
  sorting: alphabetical
---

@@

Most of the following was **copy-pasted** from published papers.

@@

The **aim** of the present study is to research the short-term effects of postnatal exposures to non-persistent \acr{edcs} on neurodevelopment and neurobehavior in childhood, and how the metabolome and the proteome might mediate these effects. We will strengthen the estimated effects by making use of the principles and criteria of triangulation [@LawlorTillingDaveySmith:2016]. To estimate these effects, we will primarily rely on the use of \acr{mtps} [@munozPopulationInterventionCausal2012; @haneuseEstimationEffectInterventions2013; @diazNonparametricCausalEffects2021] in combination with \acr{tmle} [@vanderlaanTargetedMinimumLossBased2018], to avoid relying on arbitrary parametric assumptions. We will consider both additive and multiplicative shifts of the exposures' distribution.

@@

{{< include ../../../all_phd/descriptive_stats.qmd >}}

@@

```{r}
rq1.title <- "What are the short-term effects of childhood exposure to EDCs on the outcomes?"
rq2.title <- "What are the short-term effects of childhood exposure to EDCs on metabolites and proteins?"
rq3.title <- "What are the short-term effects of childhood *exposure* to metabolites and proteins on the outcomes?"
```

@@

where $(f_{W{_t}}, f_{A_{t}}, f_{Y_{t}})$ are the non-parametric structural equations, 
and $(U_{W_{t}}, U_{A_{t}}, U_{Y_{t}})$ are the unmeasured factors contributing the confounders, exposures, and outcome, respectively. 
The corresponding causal graph (without indexes for ease of notation) is shown in @fig-causal-graph.

```{r}
#| label: fig-causal-graph
#| fig.cap: Causal graph corresponding to the structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal_graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") %>%
  ggdag::tidy_dagitty()
ggdag::ggdag(causal_graph) + ggdag::theme_dag()
```

@@

## Research question 1: `r rq1.title`

@@

## Research question 2: `r rq2.title`

We will estimate the short-term effects of childhood exposure to non-persistent \acr{edcs} on metabolites and proteins measured in childhood. We will perform an XWAS analysis with the dependent variable $m_i$ being the metabolite or protein $i$, and the independent variable $x_j$ being the levels of \acr{edc} $j$. We will assume that all associations are confounded by the same set of confounders, independently of the \acr{edc} considered.

## Research question 3: `r rq3.title`

We will estimate the short-term effects of childhood *exposure* to metabolites and proteins on the outcomes of interest measured in childhood. We will perform an XWAS analysis with the dependent variable $y$ being the outcome, and the independent variable $m_i$ being the metabolite or protein $i$. We will assume that all associations are confounded by the same set of confounders, independently of the biomarker considered.

@@

* **No unmeasured confounding**. There are no unmeasured common causes of the exposure and the subsequent outcome, which we can formalize as $Y^d \perp \!\!\! \perp A|W$. In our context, this assumption would be violated if, for example, an unmeasured variable influences both the exposure levels and the clinical outcome. We cannot guarantee that this assumption holds, and therefore limit our interpretations to statistical associations rather than causal effects.
* **Positivity**. If $(a, w)$ is within the support of $A, W$, then $(d(a,c), w)$ for additive shifts and $(d(a,k), w)$ for multiplicative shifts must also be within the support of $A,W$. In practice, this means that for any given time index and set of adjustment covariates, there is a positive probability of finding a subject with the same covariate values and a exposure level matching the shifted value. We will attempt to improve plausibility of this assumption by considering small shifts, while recognizing that this approach makes our causal effect data-adaptive.
* **Independence** of subjects. This assumption also implies **no interference**: the exposure level $a$ for a given subject does not affect the outcomes of the other subjects.
* **Consistency**. If $A = a$ for any subject, then $Y(a) = Y$, and hence the full observed set of outcomes when $A^d = A$ is simply $Y(A^d) = Y$. This means that the counterfactual outcome for a subject with its observed exposure level is the observed outcome.
* **Time-ordering**. The confounders $W$ precede the exposure $A$, which also precedes the outcome $Y$.

@@

with $\Lambda$ indicating whether each subject received the natural or shifted value of treatment. We refer to $P(\Lambda=1|a,w)$ as $\bar{\lambda}$. To implement estimation of $\bar{\lambda}$ and $\bar{Q}$ flexibly, we will use the ensemble machine learning algorithm \acr{sl} [@vanderlaanSuperLearner2007; @naimiStackedGeneralizationIntroduction2018]. In the following, we present the steps necessary to estimate $\psi_0 (A^d)$ and the associational parameter $\psi_0^{\Delta}$. We note that our bounded continuous outcome will be **scaled** to $[0,1]$, and that to handle missing covariates $W_{i,j}$, we will redefine the data to include $\delta_{ij}$, where $\delta_{ij}=1$ if the covariate was observed and $0$ otherwise.

@@

Using the empirical mean $\hat{Y} = \frac{1}{n} \sum_{i=1}^nY_i$, to estimate $\psi_0(A) = \mathbb{E}(Y)$, we obtain an estimate of the statistical association parameter $\psi_0^{\Delta} = \psi_0(A^d) - \psi_0(A)$ with
$$
\hat{\psi}^{\Delta} = \hat{\psi}_{\text{tmle}} - \bar{Y}.
$$

@@

Using `r I(cv.folds)`-fold \acr{cv}, \acr{sl} builds the best weighted combination of predictions from the algorithms listed in \@ref(tab:lmtp-specs). In case of repeated measures of the exposures, an `ID` variable will be used to make sure that the `r I(cv.folds)`-fold \acr{cv} splits will keep observations from the same individuals in the same split.

@@

- `outcome ~ chemical`
  - **Regression analysis**. We will use \acr{tmle} in combination with \acr{sl}, and a library of estimators including both simple parametric models (e.g., `glm`) and data-adaptive semi-parametric models. The models will be adjusted for relevant confounders, as identified with `dagitty`. We hypothesize the presence of residual confounding, especially due to genetic and parental \acr{se} factors, which would result in **exaggeration** of any true causal effect.
  - **Outcome negative control study**. We will try to identify outcomes in the same cohorts that would be affected by confounders that are relevant in the association of \acr{edcs} exposure to the clinical outcome of interest, but for which a biological/causal effect of the exposures is unlikely.
  - **Literature search**. For each of the `outcome ~ chemical` associations, we will perform a literature search to compare the obtained results. We will make use of PubMed, using as keywords the clinical outcome and the *common* name of each chemical, and the \acr{ctd} [@DavisWiegersJohnson:2023].
  - **Qualitative \acr{prs}**. \textcolor{red}{TODO}
- `omic ~ chemical`
  - **Regression analysis**. As above.
  - **Cross-cohort comparison**. We will compare the regression models' results between subjects of different ethnic origins. We postulate that \acr{se} factors represent the main confounders for this association. Based on the associations between \acr{se} factors (e.g., maternal education) and the omic markers, we will compare the results of `omic ~ chemical` with the expected modified associations (e.g., weaker or stronger).
  - **Literature search**. As above.
  - **Qualitative \acr{prs}**. \textcolor{red}{TODO}
- `outcome ~ omic`
  - **Regression analysis**. As above.
  - **Cross-cohort comparison**. As above.
  - **\acr{mr}**. We will make use, ideally, of a weighted allele score of genetic variants known to be robustly associated with the omic markers as an \acr{IV}. We will use methods, including sensitivity analysis, to explore the possibility of bias due to: (i) weak instruments, and (ii) violation of the exclusion restriction criteria. Based on these, we will postulate whether the obtained results might be biased towards the **null** or not. By using \acr{mr}, we assume that this approach tests *exposure* to omic markers across the whole of childhood.

## Mendelian Randomization: Assumptions

@@

## Reliable Causal Inference

@@

## Reliable Statistical Inference

@@

## Replicability

@@

## Data Checking

{{< include ../../../all_phd/app-data-checking.qmd >}}

## Checklist A: Replicability

@@

### Research question 1: `r rq1.title`

```{r}
#| fig-cap: DAG for research question 1.
#| fig-align: center
#| fig-width: 30
#| fig-height: 40
#| out-extra: "keepaspectratio=true"
dag_rq <- dags$chem_to_out %>%
  dagitty::dagitty()
dag_rq %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 12, text_col = "black") +
  ggdag::scale_adjusted() +
  ggdag::theme_dag() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))

adj_sets <- dagitty::adjustmentSets(x = dag_rq, 
                                    type = type_mas, effect = type_effect)
```

There are a total of `r length(adj_sets)` `r I(type_mas)` sufficient adjustment sets for estimating the `r I(type_effect)` effect. The first one is:

```{r}
adj_sets[[1]]
```

### Research question 2: `r rq2.title`

```{r}
#| fig-cap: DAG for research question 2.
#| fig-align: center
#| fig-width: 30
#| fig-height: 40
#| out-extra: "keepaspectratio=true"
dag_rq <- dags$chem_to_out %>%
  dagitty::dagitty()
dag_rq %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 12, text_col = "black") +
  ggdag::scale_adjusted() +
  ggdag::theme_dag() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))

adj_sets <- dagitty::adjustmentSets(x = dag_rq, 
                                    type = type_mas, effect = type_effect)
```

There are a total of `r length(adj_sets)` `r I(type_mas)` sufficient adjustment sets for estimating the `r I(type_effect)` effect. The first one is:

```{r}
adj_sets[[1]]
```

### Research question 3: `r rq3.title`

```{r}
#| fig-cap: DAG for research question 3.
#| fig-align: center
#| fig-width: 30
#| fig-height: 40
#| out-extra: "keepaspectratio=true"
dag_rq <- dags$chem_to_out %>%
  dagitty::dagitty()
dag_rq %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 12, text_col = "black") +
  ggdag::scale_adjusted() +
  ggdag::theme_dag() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))

adj_sets <- dagitty::adjustmentSets(x = dag_rq, 
                                    type = type_mas, effect = type_effect)
```

There are a total of `r length(adj_sets)` `r I(type_mas)` sufficient adjustment sets for estimating the `r I(type_effect)` effect. The first one is:

```{r}
adj_sets[[1]]
```
